Tree establishment in models of forest dynamics: a key priority for further research

### 2023-03-17 manuscript/manuscriptRW.Rmd
changes in manuscript second revision internal from Harald

---
title: "Tree establishment in models of forest dynamics: a key priority for further research"
author:
- Olalla Díaz-Yáñez,
- Yannek Käber,
- Tim Anders,
- Kristin H. Braziunas,
- Josef	Bruna,
- Samuel Fischer,
- Jessica Hetzer,
- Thomas Hickler,
- Christian	Hochauer,
- Manfred Lexer,
- Heike	Lischke,
- Mats Mahnken,
- Paola	Mairota,
- Katarina Merganicova,
- Tobias Mette,
- Xavier Morin,
- Werner Rammer,
- Simon	Scheiter,
- Daniel Scherrer,
- Harald Bugmann
date: "`r Sys.Date()`"
output:
  word_document:
    reference_docx: template_paper.docx
  bookdown::pdf_document2:
    keep_tex: yes
    number_sections: yes
    toc: no
    extra_dependencies: "subfig"
mainfont: Arial
fontsize: 12pt
linestretch: 1.5
bibliography: Regeneration.bib
link-citations: yes
nocite: |
  @bugmann1996a, @bohn2014, @fabrika2005, @mette2014, @smith2001a, @langan2017, @seidl2012a, @lischke2006, @schumacher2006, @scheiter2009, @scheller2004, @morin2021, @lasch-born2020, @hickler2012, @fischer_lessons_2016
linkcolor: blue
csl: global-change-biology.csl
editor_options:
  markdown:
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

@@

Tree establishment is a key process in forest dynamics, particularly in
the context of forest resilience and climate change. Models of forest
dynamics have been in use for more than 50 years, but an assessment of
their capacity to represent tree establishment is lacking. We assess how
well current models capture the overall abundance, species composition
and mortality of tree establishment.

Using 15 models and model variants built to capture long-term (i.e.,
decadal to centennial) forest dynamics at the stand, landscape and
global levels, we simulate tree establishment at 200 sites representing
environmental gradients across Europe. The results are evaluated against
comprehensive data from unmanaged forests.

Most of the models overestimate ingrowth levels, which is only
compensated by some models with high simulated mortality rates in the
early stages of population dynamics. Simulated species diversity of
ingrowth matches the observed ranges quite well. Models simulating
higher species diversity at the stand level do not feature higher
ingrowth diversity. The effect of light availability on ingrowth levels
is captured better than the effect of temperature and soil moisture, but
patterns are not consistent across models. The ingrowth of *Abies alba*
is captured across temperature and soil moisture gradients, whereas
ingrowth of *Fagus sylvatica* is generally underestimated. *Picea abies*
ingrowth is generally overestimated, particularly at colder sites.
Increasing complexity in the tree establishment modules of the models is
not related to higher accuracy of simulated tree establishment.
Furthermore, individual model design is more important than scale
(stand, landscape, global) and approach (empirical, process-based) to
accurately capture tree establishment.

In spite of more or less substantial mismatches between model results
and data, it is remarkable that most of the models capture the essential
features of the highly complex process of tree establishment, while not
having been parameterized with such data. We conclude that in spite of
all the uncertainties, a lot can be gained by a focus on the modeling of
establishment processes in these models, this will make their long-term
forest dynamics projections under changing environmental conditions much
more robust.

@@

Forests provide a wide range of products and services of vital
importance to humankind [@foresteurope2020]. Under the pressure of
climate change and increasing disturbances, it is becoming ever more
important to understand how forest structure, composition and function
will change, and to evaluate forest capacity to adapt or be resilient
after disturbance [@lindner2010; @seidl2022]. A wide range of models of
forest dynamics were developed over the past decades considering the
impacts of climate on forest dynamics [@bugmann2022]. From these
studies, it is evident that while we have a reasonably good
understanding how to model tree growth [@vanclay1997; @bugmann1996], and
substantial efforts have been dedicated to improving the depiction of
tree mortality [@bugmann2019]. However, tree establishment remains a
much less studied process that is often represented rather coarsely in
these models [@walck2011; @leishman1992], which is a clear research gap
in the context of climate-induced forest disturbances and forest
resilience.

Tree establishment arises from multiple processes such as reproduction,
fruit maturation, seed production, dispersal, germination, juvenile
growth and survival [@price2001; @vacchiano2018]. All these processes
are difficult to assess, and some of them are scarcely understood and
thus appear highly stochastic [@bogdziewicz2021]. Identifying the
appropriate level of complexity for the formulation of the key factors
that are leading to successful tree establishment is challenging.
Currently, tree establishment processes in dynamic forest models are
handled in a multitude of ways [@konig_tree_2022; @bugmann2022]: a) from
entirely ignoring it [as done in classical forest growth models, e.g.,
Pretzsch et al. -@pretzsch2002], b) across the use of a few simple
environmental filters, as done in most forest "gap" models [@shuman2011]
and Dynamic Global Vegetation Models [e.g., @hickler2012; @smith2001],
c) to complex approaches that incorporate local feedback from the
canopy, multiple ecological processes and often also small time steps
[e.g., @seidl2012; @wehrli2006], or d) field-based statistical
parameterizations, which however are not easy to extrapolate in space
and time [e.g., @ribbens1994].

Overall, models are needed to (1) synthesize existing empirical data and
explore their consequences, (2) assess future tree establishment e.g.,
in the context of global change scenarios and, (3) identify most
important processes needed to define relevant ecological processes.
Given the current strategies that are used in models of forest dynamics
to represent tree establishment, model behavior often faces problems
that can have strong impacts on simulated forest dynamics, such as very
high levels of tree establishment, which then necessitate excess
mortality at early stages of tree life to simulate correct stand
structure and composition. Also, inadequately high species diversity in
tree establishment may be simulated. This is characteristic of
"classical" forest gap models [@gutiérrez2016], at least as long as the
simulations comprise a multitude of species. Other models use
calibration against local canopy-level data to define appropriate
ingrowth levels, which is likely to strongly hamper the general
applicability of these models e.g., under scenarios of climate change.
Furthermore, correctly capturing the species composition in tree
establishment as a function of the presence of seed trees in the canopy
is often a particular challenge, potentially leading to unrealistic
successional drift in the model, which must be corrected by factors that
are hard to parameterize [e.g., @lischke2006]. A related issue is the
excessive reduction of species diversity due to positive feedback
effects such that eventually just single-species stands remain. This is
sometimes corrected by the incorporation of a low level of seed influx
of all species at all times [@schumacher2006] or by restricting the
number of seeds per species in the seedbank [@lischke2006]. However,
simulated species composition is usually exceedingly sensitive to that
level, whereas the parameters of such functions are poorly constrained
by field data. Lastly, there is often a problem with insufficient
observational constraints on parameter values for models that start from
very small tree sizes (e.g., 10 cm of height or even seeds) and track
tree development in a process-oriented manner by considering a multitude
of ecological influences, rather than emphasizing tree establishment
into a larger size class [e.g., @wehrli2006].

Thus, a focus on the modeling of tree establishment processes is sorely
needed and overdue [@price2001; @walck2011] if we are to make reliable
projections of future forest dynamics, i.e., when the models need to be
operated in extrapolation mode, but also from a fundamental ecological
point of view increasing systems understanding. Also in the field-based
ecology, tree establishment has not received a lot of attention
[@hanbury-brown2022], although many sub-processes have been studied in
detail. Yet, little data is available that covers all the processes
within one species along environmental gradients, let alone for a vast
suite of species. Thus, the available data are fragmented, which
constitutes a major problem for model building [@clark1999]. On the one
hand, monitoring on permanent sample plots (such as National Forest
Inventories) often measures tree establishment [@zell2019], however with
a design that captures the rate after the stand initiation phase
[@hallsby2015]. Data from permanent plots, although available across
regions, are highly heterogeneous, rendering their use difficult in a
modeling context [@käber2021]. On the other hand, targeted studies to
measure tree establishment on experimental sites are often limited in
spatial extent, species studied, or the sub-set of processes that are
targeted [@berdanier2016; @collet2006].

In the present study, for the first time, a large number of forest
models commonly used to assess forest dynamics under climate change are
evaluated against a continental-scale, multi-species harmonized dataset
on tree establishment [@käber2022]. This is a challenging experiment for
these models as they do also have to estimate the forest stand above the
tree establishment. By tree establishment, we refer to the passing of
trees across a specific diameter threshold ("ingrowth"). We included
models that are based on a range of "philosophies", from models
operating at the stand to the global scale as well as the range of
models from empirically derived to "process-based" [e.g., @bugmann1996;
@fabrika2012; @lexer2001; @reyer2014]. Because of the large variability
in tree establishment patterns in nature and the large number of factors
driving this process --- including some that are not incorporated in
most models such as deer browsing --- we do not aim at a detailed
statistical evaluation of each model. Instead, we aim at evaluating the
general ingrowth patterns and magnitudes simulated by the models, and
benchmark the simulated regeneration niche of multiple species against
empirical data along a wide environmental gradient of temperature,
moisture, and light availability.

More specifically, we aimed to answer the following questions: (1) Are
models of forest dynamics capturing accurately tree ingrowth levels,
initial tree diversity and mortality in tree establishment? (2) Do model
traits explain differences in model performance? (3) How well do models
capture total ingrowth and individual species regeneration niches across
environmental gradients of light availability, temperature, and soil
moisture?

@@

Fifteen models of forest dynamics (including two models featuring
variants) were used to assess simulated tree establishment. They
simulate forest dynamics at the stand, landscape or global scale (Table
\@ref(tab:tableModels)). The approaches used for model construction and
their origin differ strongly, with most of the models featuring a
largely "process-based" approach, whereas two models are based on
formulations derived from the statistical analysis of inventory data
(SIBYLA and xComp). Some of the models largely rely on the original
approach underlying forest gap models (e.g., ForClim 1 and LandClim),
whereas some are based on plant ecophysiological processes (e.g.,
FORMIND and iLand). The different approaches underlying the models have
strongly influenced the formulation of tree establishment.

@@

The starting point for the models establishment differs as well, ranging
from seedbank, seed or seedling to sapling (i.e., trees much larger than
10 cm height, often at least 2-3 m tall). Models that start from seed
need to include a larger number of ecological processes such as
germination and survival of young seedlings, whereas models that start
from saplings have to aggregate via parameterizations several ecological
processes because they are not treated explicitly. This reduces model
complexity but comes at the cost of blurring process representation. In
the models used in this study the overall complexity in the
establishment modules varies considerably. Following @bugmann2022, we
can classify seven models (iLand, PICUS, LandClim, ForCEEPS, LPJ-GUESS,
ForClim and TreeMig) as having rather high complexity in their
establishment modules (mean establishment formulation complexity
complexity across all processes \>0, Table \@ref(tab:tableModels)).

Two models, ForClim and ForCEEPS, were used here with two alternative
variants. ForClim variant 1 [@bugmann1996] is based on a recruitment
module that adheres closely to the concept introduced by @botkin1972,
whereas ForClim variant 11 is adopting a different approach where
individual species properties and its relation to the environment are
more important [@huber2020]. These two model variants allow us to
evaluate a more process-based and complex module (variant 11) against a
simple module (variant 1), while the rest of the model structure is
identical. Similarly, the two ForCEEPS variants allow us to isolate the
importance of the canopy feedback (i.e., simulated actual composition
and relative abundance of species in the plot) via seed trees for the
quantity and quality of simulated ingrowth, as one variant includes this
feedback whereas the other does not.

@@

Ingrowth data covering a wide range of environmental conditions are hard
to obtain, and this is one of the reasons why most models of forest
dynamics have never been confronted with a dataset covering such
gradients over a large number of sites, to evaluate how well ingrowth is
captured. The observations used here are derived from a novel and
unprecedented network of sites in forest reserves that represent the
range of environmental gradients in temperature and precipitation in
central Europe as compiled in the framework of the EuFoRIa network
[@euforia2019] (Figure \@ref(fig:map)). These forests are unmanged for
at least 10 years prior to their designation as reserves, and most of
them provide time series of natural forest dynamics over multiple
decades and up to ca. 80 years. The census periods range from 3 to 37
years, with an average of 14 years. The data provide information at the
tree level, thus allowing for sequential comparison such as
individual-tree ingrowth and death between the measurements.

We selected 200 sites from this network as the benchmarking dataset for
the simulation. The selection of sites is representative for the
environmental variation of the data. This was achieved by applying
k-means clustering to define 200 clusters of plots from the original set
of 869 plots along the environmental dimensions of temperature, water
balance, soil quality and exposition (i.e., slope and aspect). Each of
these sites featured at least two consecutive measurements. Ingrowth
thresholds for the sites differed between 0 and 10 cm, we defined two
data sets where one included all sites with a diameter threshold of 7 cm
or lower and the other included another 35 sites with diameter
thresholds between 7 and 10 cm. Thus we focused on two data sets with
the harmonized diameter thresholds of 7 cm and 10 cm.

Plot size ranged from 0.02 ha to 5.52 ha, and the data were further
processed and aggregated following @käber2022 to provide ingrowth rates
per ha and per decade. The observations used in this study featured
30,900 established trees, and ingrowth rates per site, sample and
decade ranged from zero to 1246 trees per ha, with a mean of 56 trees
per ha. Adult species composition was also available for the simulated
species. For more details on this unique dataset and the detailed
establishment information, cf. @käber2022.

@@

The overarching goal of the experiments was to assess the tree
establishment as it arises from empirical data against its
representation in a wide range of models of forest dynamics. We define
the tree establishment as the passing of a breast height diameter
threshold of 7 or 10 cm, respectively (synonym: ingrowth). To this end,
each modeling group was provided with a detailed protocol (cf.
Supplementary Material 1, henceforth "SM") with instructions how to
perform the simulations, providing input variables on climate and soil
conditions, and the expected output variables. Neither were further site
information (except for the data specified below) nor any data on tree
ingrowth provided prior to the simulation. That is, the models were run
in "blind flight" mode.

The input variables were collected and aggregated to be adapted to the
needs of the different models. Time series of climatic variables from
1981 to 2018 were provided in hourly [Era5-land data, Muñoz Sabater
-@muñozsabater2019], daily and monthly resolution [CHELSA data, Karger
et al. -@karger2021]. Some variables required for some models such as
relative humidity or vapor pressure deficit had to be calculated from
these variables. The final strategy how to use the climate data was
different in each model, based on the approach that best suited the
model (Table \@ref(tab:tableModels)). Soil quality data was provided as
a continuous value between 1 and 5 [Soilgrids dataset, Hengl et al.
-@hengl2017]. The protocol also provided the elevation, slope and aspect
for each of the 200 sites [@asterscienceteam2019], but no other spatial
information such as coordinates was provided, with the exception of
iLand and aDGVM2, which used blurred coordinates to derive highly
detailed soil data.

@@

The simulations were run for mixed species forests (not multiple
single-species simulations) using mixtures of eleven species or genera
for which ingrowth data of sufficient quality were available from
EuFoRIa: *Fagus sylvatica, Picea abies, Abies alba, Carpinus betulus,
Tilia cordata, Acer pseudoplatanus, Betula spp., Fraxinus excelsior,
Quercus spp., Alnus glutinosa, Pinus sylvestris.* The same set of
species were used in all the 200 sites. Two models included their
standard set of species for the simulations which is much larger (i.e.,
ForClim 1, ForClim 11, TreeMig). In three models, fewer than these
eleven species were simulated (4C, xComp, LPJ-GUESS) (Table
\@ref(tab:tableModels)). In 4C only *Fagus sylvatica*, *Picea abies*,
*Betula* spp., *Quercus* spp. and *Pinus sylvestris* are parameterised.
XComp simulations did not consider *Fraxinus excelsior* due to a
depreciated species parameterisation. In LPJ-GUESS, *Acer
pseudoplatanus* and *Alnus glutinosa* are not parameterized and
therefore these species could not be included in the simulation.
Finally, the model aDGVM2 does not simulate individual species. Rather,
community assembly processes and trait filtering generate plant
communities that are adjusted to the biotic and abiotic conditions and
plants can be classified into different ecological strategies based on
their trait values in a post-processing step.

@@

Three models did not provide results from all the simulated samples or
sites to avoid unrealistic results, as follows. In the model 4C, a
threshold of a maximum basal area of 90 $m^2 ha^{-1}$ was used to avoid
unrealistic stand basal area data, and therefore not all the samples and
sites were reported. The reason for this is that 4C is not suitable for
long-term simulations without management, due to misrepresentations in
density-dependent mortality processes in long-term simulations and
assumptions of tree geometry that lead to unrealistic single tree
dimensions of very old individuals. LPJ-GUESS had 2% of the sample
outputs without tree establishment and were considered in the data as
zero stand basal area and zero ingrowth for all the tree species
simulated and one site produced grasslands and was not included in the
results.

@@

The simulation results were analyzed regarding the ingrowth levels
(i.e., ingrowth number per unit time and space), species diversity of
the ingrowth, and ingrowth mortality. We evaluated the species diversity
across models and in relation to the observed data by calculating the
Shannon index based on the relative proportion of the species in terms
of basal area. The Shannon index was calculated for the establishment
($H_{R_{n}}$) (Equation \ref{eqpiShannonR}) and at the stand level
($H_{S_{n,s}}$)(Equation \ref{eqpiShannon}). The higher the value of the
index, the higher the species diversity at a particular site and sample.

>>>>

---
title: "Tree establishment in models of forest dynamics: a key priority for further research"
author:
- Olalla Díaz-Yáñez,
- Yannek Käber,
- Tim Anders,
- Kristin H. Braziunas,
- Josef	Bruna,
- Samuel Fischer,
- Jessica Hetzer,
- Thomas Hickler,
- Christian	Hochauer,
- Manfred Lexer,
- Heike	Lischke,
- Mats Mahnken,
- Paola	Mairota,
- Katarina Merganicova,
- Tobias Mette,
- Xavier Morin,
- Werner Rammer,
- Simon	Scheiter,
- Daniel Scherrer,
- Harald Bugmann
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    keep_tex: yes
    number_sections: yes
    toc: no
    extra_dependencies: "subfig"
  word_document:
    reference_docx: template_paper.docx
mainfont: Arial
fontsize: 12pt
linestretch: 1.5
bibliography: Regeneration.bib
link-citations: yes
nocite: |
  @bugmann1996a, @bohn2014, @fabrika2005, @mette2014, @smith2001a, @langan2017, @seidl2012a, @lischke2006, @schumacher2006, @scheiter2009, @scheller2004, @morin2021, @lasch-born2020, @hickler2012, @fischer_lessons_2016
linkcolor: blue
csl: global-change-biology.csl
editor_options:
  markdown:
    wrap: 72
header-includes:
- \usepackage[left]{lineno}
- \linenumbers
---

@@

Tree establishment is a key process in forest dynamics, particularly in
the context of forest resilience and climate change. Models are pivotal
for assessing long term (i.e., decadal to centennial) forest dynamics,
and they have been in use for more than 50 years. However, an assessment
of their ability to accurately represent tree establishment is lacking.
We assess how well current models capture the overall abundance, species
composition and mortality of tree establishment.

Using 15 models and model variants built to capture long-term forest
dynamics at the stand, landscape and global levels, we simulate tree
establishment at 200 sites representing large environmental gradients
across Europe. The results are evaluated against comprehensive data from
unmanaged forests.

Most of the models overestimate ingrowth levels, which is only
compensated by some models with high simulated mortality rates in the
early stages of population dynamics. Simulated species diversity of
ingrowth matches the observed ranges quite well. Models simulating
higher species diversity at the stand level do not feature higher
ingrowth diversity. The effect of light availability on ingrowth levels
is captured better than the effect of temperature and soil moisture, but
patterns are not consistent across models. The ingrowth of *Abies alba*
is captured well by most models across temperature and soil moisture
gradients, whereas ingrowth of *Fagus sylvatica* is generally
underestimated. *Picea abies* ingrowth is generally overestimated,
particularly at colder sites. Increasing complexity in the tree
establishment modules of the models is not related to higher accuracy of
simulated tree establishment. Furthermore, individual model design is
more important than scale (stand, landscape, global) and approach
(empirical, process-based) to accurately capture tree establishment.

Despite sometimes considerable mismatches between simulation results and
data, it is remarkable that most of the models capture the essential
features of the highly complex process of tree establishment, while not
having been parameterized with such data. We conclude that in spite of
all the uncertainties, a lot can be gained by a focus on the modeling of
establishment processes in these models. This has the potential to
render long-term projections of forest dynamics under changing
environmental conditions much more robust.

@@

Forests provide a wide range of products and services of vital
importance to humankind [@foresteurope2020]. Under the pressure of
climate change and increasing disturbances, it is becoming ever more
important to understand how forest structure, composition and function
will change, and to evaluate forest capacity to adapt to or be resilient
after disturbance [@lindner2010; @seidl2022]. A wide range of models of
forest dynamics were developed over the past decades considering the
impacts of climate on forest dynamics [@bugmann2022]. From these
studies, it is evident that we have a reasonably good understanding how
to model tree growth [@vanclay1997; @bugmann1996], and substantial
efforts have been dedicated to improving the representation of tree
mortality [@bugmann2019; @cailleret2017]. In contrast, tree
establishment is a much less studied process that is often represented
rather coarsely in these models [@walck2011; @leishman1992; @price2001],
which is a clear research gap in the context of climate-induced forest
disturbances and forest resilience.

Tree establishment arises from multiple processes such as pollination,
fruit maturation, seed production, dispersal, germination, juvenile
growth and survival [@price2001; @vacchiano2018]. All these processes
are difficult to assess, and some of them are scarcely understood and
thus appear highly stochastic [@bogdziewicz2021]. Identifying the
appropriate level of complexity for the mathematical formulation of the
key factors that are leading to successful tree establishment is
challenging. Currently, tree establishment processes in dynamic forest
models are handled in a multitude of ways [@konig_tree_2022;
@bugmann2022]: from a) entirely ignoring it [as done in classical forest
growth models, e.g., Pretzsch et al. -@pretzsch2002], across b) the use
of a few simple environmental filters, as done in most forest "gap"
models [@shuman2011] and Dynamic Global Vegetation Models [e.g.,
@hickler2012; @smith2001], to c) complex approaches that incorporate
local feedback from the canopy, multiple ecological processes and often
also small time steps [e.g., @seidl2012; @wehrli2006], or d) field-based
statistical parameterizations, which however are not easy to extrapolate
in space and time [e.g., @ribbens1994].

Overall, models are needed to (1) synthesize existing empirical data and
explore their consequences, (2) assess future tree establishment e.g. in
the context of global change scenarios, and (3) identify the most
important processes that are shaping ecological patterns. Given the
current strategies that are used in models of forest dynamics to
represent tree establishment, model behavior often is prone to problems,
such as very high levels of tree establishment that necessitate excess
mortality at early stages of tree life to simulate correct stand
structure and composition. Also, inadequately high species diversity in
tree establishment may be simulated, which is characteristic of
"classical" forest gap models [@gutiérrez2016], at least as long as the
simulations comprise a multitude of species. Some models use calibration
against local canopy-level data to define appropriate ingrowth levels,
which is likely to strongly hamper the general applicability of these
models e.g. under scenarios of climate change. Furthermore, correctly
capturing the species composition in tree establishment as a function of
the presence of seed trees in the canopy is often a particular
challenge, potentially leading to unrealistic successional drift in the
model, which must be corrected by factors that are hard to parameterize
[e.g., @lischke2006]. A related issue is the excessive reduction of
species diversity due to positive feedback effects such that eventually
just single-species stands remain. This is sometimes corrected by the
incorporation of a low level of seed influx of all species at all times
[@schumacher2006] or by restricting the number of seeds per species in
the seedbank [@lischke2006]. However, simulated species composition is
usually exceedingly sensitive to that level, whereas the parameters of
such functions are poorly constrained by field data. Lastly, there is
often a problem with insufficient observational constraints on parameter
values for models that start from very small tree sizes (e.g., 10 cm of
height or even seeds) and track tree development in a process-oriented
manner by considering a multitude of ecological influences, rather than
emphasizing tree establishment into a larger size class [e.g.,
@wehrli2006].

Thus, a focus on the modeling of tree establishment processes is sorely
needed and overdue [@price2001; @walck2011] if we are to make reliable
projections of future forest dynamics, i.e., when the models need to be
operated in extrapolation mode, as well as from a fundamental ecological
point of view for increasing systems understanding. Also in empirical
ecology, tree establishment has not received a lot of attention
[@hanbury-brown2022], although many sub-processes have been studied in
detail. Yet, little data is available that covers all the processes
within one species along environmental gradients, let alone for a vast
suite of species. Thus, the available data are fragmented, which
constitutes a major problem for model building [@clark1999]. On the one
hand, monitoring on permanent sample plots (such as National Forest
Inventories) often measures tree establishment [@zell2019], however with
a design that captures the rate after the stand initiation phase
[@hallsby2015]. Data from permanent plots, although available across
regions, are highly heterogeneous, rendering their use difficult in a
modeling context [@käber2021]. On the other hand, targeted studies to
measure tree establishment on experimental sites are often limited in
spatial extent, species studied, or the sub-set of processes that are
targeted [@berdanier2016; @collet2006].

In the present study, for the first time, a large number of forest
models commonly used to assess forest dynamics under climate change are
evaluated against a continental-scale, multi-species harmonized dataset
on tree establishment [@käber2022]. By tree establishment, we refer to
the passing of trees across a specific diameter threshold ("ingrowth").
We included models that are based on a range of "philosophies", from
models operating at the stand to the global scale as well as the range
of models from empirically derived to "process-based" [e.g.,
@bugmann1996; @fabrika2012; @lexer2001; @reyer2014]. Because of the
large variability in tree establishment patterns in nature and the large
number of factors driving this process --- including some that are not
incorporated in most models such as deer browsing --- we do not aim at a
detailed statistical evaluation of each model. Instead, we aim at
evaluating the general ingrowth patterns and magnitudes simulated by the
models, and benchmark the simulated regeneration niche of multiple
species against empirical data along a wide environmental gradient of
temperature, moisture, and light availability.

More specifically, we aim to answer the following questions: (1) Are
models of forest dynamics capturing accurately tree ingrowth levels,
initial tree species diversity and mortality in the establishment? (2)
Do model traits explain differences in model performance? (3) How well
do the models capture total ingrowth and individual species regeneration
niches across environmental gradients of light availability,
temperature, and soil moisture?

@@

Fifteen models of forest dynamics (including two models featuring
variants) were used to assess simulated tree establishment. They
simulate forest dynamics at the stand, landscape or global scale (Table
\@ref(tab:tableModels)). The approaches used for model construction and
their origin differ strongly, with most of the models featuring a
largely "process-based" approach, whereas two models are based on
formulations derived from the statistical analysis of inventory data
(SIBYLA and xComp). Some of the models largely rely on the original
approach underlying forest gap models (e.g., ForClim 1 and LandClim),
whereas some are based on plant ecophysiological processes (e.g.,
FORMIND and iLand). The different approaches underlying the models have
strongly influenced the formulation of tree establishment processes.

@@

The starting point for tree establishment in the models differs as well,
ranging from seedbank, seed or seedling to sapling (i.e., trees much
larger than 10 cm height, often ca. 2-3 m tall). Models that start from
seed need to include a larger number of ecological processes such as
germination and survival of young seedlings, whereas models that start
from saplings have to aggregate via parameterizations several ecological
processes because they are not treated explicitly. This later approach
reduces model complexity but comes at the cost of blurring process
representation.

In the models used in this study, the overall complexity in the
establishment modules varies considerably. Following @bugmann2022, we
can classify seven models (iLand, PICUS, LandClim, ForCEEPS, LPJ-GUESS,
ForClim and TreeMig) as having rather high complexity in their
establishment modules (mean establishment formulation complexity
complexity across all processes \>0, Table \@ref(tab:tableModels)). Two
models, ForClim and ForCEEPS, were used here with two alternative
variants of establishment. ForClim variant 1 [@bugmann1996] is based on
a recruitment module that adheres closely to the concept introduced by
@botkin1972, whereas ForClim variant 11 is adopting a different approach
where individual species properties and their relation to the
environment are important [@huber2020]. These two model variants allow
us to evaluate a more process-based and complex module (variant 11)
against a simple module (variant 1), while the rest of the model
structure is identical. Similarly, the two ForCEEPS variants allow us to
isolate the importance of the canopy feedback (i.e., simulated actual
composition and relative abundance of species in the plot) via seed
trees for the quantity and quality of simulated ingrowth, as one variant
includes this feedback whereas the other does not.

@@

Ingrowth data covering a wide range of environmental conditions are hard
to obtain, and this is one of the reasons why most models of forest
dynamics have never been confronted with a dataset covering such
gradients over a large number of sites, to evaluate how well ingrowth is
captured. The observations used here are derived from a novel and
unprecedented network of sites in forest reserves that represent the
range of environmental gradients in temperature and precipitation in
central Europe as compiled in the framework of the EuFoRIa network
[@euforia2019] (Figure \@ref(fig:map)). These forests have been
unmanaged for at least 10 years prior to their designation as reserves,
and most of them provide time series of natural forest dynamics over
multiple decades and up to ca. 80 years. The census periods range from 3
to 37 years, with an average of 14 years. The data provide information
at the tree level, thus allowing for sequential comparison such as
individual-tree ingrowth and death between the measurements.

We selected 200 sites from this network as the benchmarking dataset for
the simulation. The selection of sites is representative for the
environmental variation of the data. This was achieved by applying
k-means clustering to define 200 clusters of plots from the original set
of 869 plots along the environmental dimensions of temperature, water
balance, soil quality and exposition (i.e., slope and aspect). Each of
these sites featured at least two consecutive measurements. Ingrowth
thresholds for the sites differed between 0 and 10 cm. For our study, we
defined two data sets where one included 165 sites with a diameter
threshold of 7 cm or lower, and the other included another 35 sites with
diameter thresholds between 7 and 10 cm. Thus, we focused on two data
sets with harmonized diameter thresholds of 7 cm and 10 cm,
respectively.

Plot size ranged from 0.02 ha to 5.52 ha, and the data were further
processed and aggregated following @käber2022 to provide ingrowth rates
per ha and per decade. The observations used in this study featured
30,900 established trees. Ingrowth rates per site, sample and
decade ranged from zero to 1246 trees per ha, with a mean of 56 trees
per ha. Adult species composition was also available for the simulated
species. For more details on this unique dataset and the detailed
establishment information, cf. @käber2022.

@@

The overarching goal of the experiments was to assess the tree
establishment as it arises from empirical data against its
representation in a wide range of models of forest dynamics. We define
the tree establishment as the passing of a breast height diameter
threshold of 7 or 10 cm, respectively (synonym: ingrowth). To this end,
each modeling group was provided with a detailed protocol (cf.
Supplementary Material 1, henceforth "SM") with instructions how to
perform the simulations, providing input variables on climate and soil
conditions, and the expected output variables. Neither were further site
information (except for the data specified below) nor any data on tree
ingrowth or forest stand features provided prior to the simulation. That
is, the models were run in "blind flight" mode.

The input variables were collected and aggregated to be adapted to the
needs of the different models. Time series of climatic variables from
1981 to 2018 were provided in hourly [Era5-land data, Muñoz Sabater
-@muñozsabater2019], daily and monthly resolution [CHELSA data, Karger
et al. -@karger2021]. Some variables required for some models, such as
relative humidity or vapor pressure deficit, were calculated from these
variables. The final strategy how to use the climate data was different
in each model, based on the approach that best suited the model (Table
\@ref(tab:tableModels)). Soil quality data were provided as continuous
values between 1 and 5 [Soilgrids dataset, Hengl et al. -@hengl2017].
The protocol also provided the elevation, slope and aspect for each of
the 200 sites [@asterscienceteam2019], but no other spatial information
such as coordinates, with the exception of iLand and aDGVM2, which
required blurred coordinates to derive highly detailed soil data.

@@

The simulations were run for mixed species forests (not multiple
single-species simulations) using mixtures of eleven species or genera
for which ingrowth data of sufficient quality were available from
EuFoRIa: *Fagus sylvatica, Picea abies, Abies alba, Carpinus betulus,
Tilia cordata, Acer pseudoplatanus, Betula spp., Fraxinus excelsior,
Quercus spp., Alnus glutinosa, Pinus sylvestris.* The same set of
species was used at all 200 sites. Two models included their standard
set of species for the simulations, which is much larger (i.e., ForClim
1, ForClim 11, TreeMig). In three models, fewer than these eleven
species were simulated (4C, xComp, LPJ-GUESS) (Table
\@ref(tab:tableModels)). In 4C only *Fagus sylvatica*, *Picea abies*,
*Betula* spp., *Quercus* spp. and *Pinus sylvestris* are parameterised.
XComp simulations did not consider *Fraxinus excelsior* due to a
depreciated species parameterisation. In LPJ-GUESS, *Acer
pseudoplatanus* and *Alnus glutinosa* are not parameterized and
therefore these species could not be included in the simulation.
Finally, the model aDGVM2 does not simulate individual species. Rather,
community assembly processes and trait filtering generate plant
communities that are adjusted to the biotic and abiotic conditions, and
the simulated plants can be classified into ecological strategies based
on their trait values in a post-processing step.

@@

Three models did not provide results from all the simulated samples or
sites to avoid unrealistic results, as follows. In the model 4C, a
threshold of a maximum basal area of 90 $m^2 ha^{-1}$ was used to avoid
unrealistic stand basal area data, and therefore not all the samples and
sites were reported. The reason for this is that 4C is not suitable for
long-term simulations without management, due to misrepresentations in
density-dependent mortality processes in long-term simulations and
assumptions of tree geometry that lead to unrealistic single tree
dimensions of very old individuals. LPJ-GUESS had 2% of the sample
outputs without tree establishment, and these were considered as zero
stand basal area and zero ingrowth for all the tree species simulated;
one site produced grassland rather than a forest and was not included in
the results.

@@

The simulation results were analyzed regarding (1) ingrowth levels
(i.e., ingrowth number per unit time and space), (2) species diversity
of the ingrowth, and (3) ingrowth mortality. We evaluated the species
diversity across models and in relation to the observed data by
calculating the Shannon index based on the relative proportion of the
species in terms of basal area. The Shannon index was calculated for the
establishment ($H_{R_{n}}$) (Equation \ref{eqpiShannonR}) and at the
stand level ($H_{S_{n,s}}$)(Equation \ref{eqpiShannon}). The higher the
value of the index, the higher the species diversity at a particular
site and sample.

```{=tex}
\begin{equation}
H_{R_{n}} = - \sum_{i = 1}^{s}{p_{R_{n,i}}\ln(p_{R_{n,i}})}\label{eqpiShannonR}
\end{equation}
```
```{=tex}
\begin{equation}
p_{R_{n,i}} = \frac{r_{BA_{n,i}}}{TotalR_{BA_{n}}} \label{eqpiShannonR2}
\end{equation}
```
```{=tex}
\begin{equation}
H_{S_{n}} = - \sum_{i = 1}^{s}{p_{S_{n,i}}\ln(p_{S_{n,i}})}\label{eqpiShannon}
\end{equation}
```
```{=tex}
\begin{equation}
p_{S_{n,i}} = \frac{BA_{n,i}}{Total_{BA_{n}}} \label{eqpiShannon2}
\end{equation}
```
where $s$ is the total number of species present and that have a basal
area larger than zero in sample $n$; $p_{n,r,i}$ is the proportion of
the species $i$ in sample $n$ calculated as the ingrowth basal area
($r_{BA_{n,i}}$) for that species $i$ relative to the total recruited
basal area ($TotalR_{BA_{n}}$) of the sample $n$; $p_{S_{n,i}}$ is the
proportion of the species $i$ calculated as the basal area of all trees
($BA_{n,i}$) of that species $i$ relative to the total basal area
($TotalR_{BA_{n}}$) of the sample $n$.

Mortality in tree establishment was assessed based on the ratio of
ingrowth between the 7 and 10 cm diameter thresholds. We used the
Reineke self-thinning rule [@reineke1933; @pretzsch2003] as a reference
to estimate whether the ratio of ingrowth between the 7 and 10 cm
diameter thresholds were above or below the expected theoretical rate
between these two diameters. The Reineke self-thinning rule is usually
calculated from even-aged, single species stands and assumes a fixed
relationship between the number of stems and the quadratic mean diameter
in fully stocked pure stands. The value used in our comparisons was 1.77
(i.e., we expect stem numbers at 7 cm to be 77% higher than at 10 cm),
calculated using Equation \ref{eqReineke}.
